---
title: "Infectious Disease Modelling"
subtitle: "An introduction for survival analysts"
author: "Kylie Ainslie"
date: today
format:
  revealjs:
    theme: simple
    slide-number: c/t
    incremental: false
    chalkboard: true
    transition: slide
    fig-align: center
    fig-width: 8
    fig-height: 4.2
    code-fold: false
    code-line-numbers: false
    smaller: false
    width: 1280
    height: 720
    margin: 0.06
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false
    dev: "svglite"
    fig.align: "center"
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(deSolve)
})
theme_set(theme_minimal(base_size = 16) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold")))
pal <- c(S = "#1f77b4", E = "#9467bd", I = "#d62728", R = "#2ca02c", D = "#7f7f7f")
```

## What is infectious disease modelling? {.smaller}

:::: {.columns}
::: {.column width="55%"}
> The quantitative study of how infectious diseases move through populations.

Mathematical and statistical models of:

- **states** an individual can be in — susceptible, infectious, recovered, …
- **transitions** between those states
- **inference** of the underlying rates from incomplete surveillance data

The unit of analysis is a *population*, not a single patient.
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
sir_intro <- function(t, y, p) {
  with(as.list(c(y, p)), {
    dS <- -beta * S * I / N
    dI <-  beta * S * I / N - gamma * I
    dR <-  gamma * I
    list(c(dS, dI, dR))
  })
}
out <- as.data.frame(ode(c(S = 0.999, I = 0.001, R = 0),
                         seq(0, 120, 0.5), sir_intro,
                         parms = c(beta = 0.4, gamma = 0.15, N = 1)))
out |>
  pivot_longer(-time, names_to = "state") |>
  ggplot(aes(time, value, colour = state)) +
  geom_line(linewidth = 1.6) +
  scale_colour_manual(values = pal) +
  annotate("text", x = 40, y = 0.32, label = "hidden states\n(what we model)",
           size = 4.2, fontface = "italic", colour = "grey25") +
  annotate("curve", x = 38, y = 0.27, xend = 32, yend = 0.18,
           arrow = arrow(length = unit(0.18, "cm")),
           curvature = 0.25, colour = "grey25") +
  labs(x = "day", y = "fraction of population",
       colour = NULL, title = "An outbreak, abstracted") +
  theme(legend.position = "bottom")
```
:::
::::

## What does it try to do? {.smaller}

```{mermaid}
%%| fig-width: 11
flowchart LR
  A["<b>Describe</b><br/>What is the epidemic<br/>doing right now?<br/><br/><i>estimate FoI, R<sub>t</sub>,<br/>attack rate, IFR</i>"]:::desc
  B["<b>Predict</b><br/>What will it do next?<br/><br/><i>forecast cases,<br/>hospitalisations, deaths</i>"]:::pred
  C["<b>Decide</b><br/>What should we<br/>do about it?<br/><br/><i>evaluate NPIs,<br/>vaccination, quarantine</i>"]:::dec
  classDef desc fill:#cfe2f3,stroke:#1f77b4,color:#0b3d66,stroke-width:2px
  classDef pred fill:#fff2cc,stroke:#b58900,color:#5a4500,stroke-width:2px
  classDef dec  fill:#d9f1d9,stroke:#2ca02c,color:#1b4a1b,stroke-width:2px
```

::: {.fragment .fade-in}
All three answered by **inferring rates from incomplete data** — the survival analyst's daily work.
:::

## How is that similar to survival analysis? {.smaller}

:::: {.columns}
::: {.column width="55%"}
**Both are time-to-event sciences.** They share the central object — the **hazard function** — and almost the entire toolkit:

| Survival analysis | ID modelling |
|---|---|
| Hazard $h(t)$ | Force of infection $\lambda(t)$ |
| $S(t) = e^{-\int h}$ | Susceptible fraction $= e^{-\int \lambda}$ |
| Multistate models | SIR / SEIR compartments |
| Sojourn times | Latent / infectious periods |
| Competing risks | Recovery vs death |
| Frailty | Heterogeneous susceptibility |
| Right-censoring | Reporting delay |

::: {.fragment .fade-in}
> Same mathematics, two communities — separated mostly by vocabulary.
:::
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
t <- seq(0, 10, by = 0.05)
df <- tibble(t, value = exp(-0.25 * t))
ggplot(df, aes(t, value)) +
  geom_line(linewidth = 1.8, colour = "black") +
  annotate("text", x = 1.4, y = 0.85,
           label = "S(t) — survival",
           hjust = 0, colour = "#1f77b4", size = 5.5, fontface = "bold") +
  annotate("curve", x = 1.3, y = 0.85, xend = 0.6, yend = 0.86,
           arrow = arrow(length = unit(0.18, "cm")),
           curvature = -0.2, colour = "#1f77b4") +
  annotate("text", x = 4.5, y = 0.45,
           label = "S(t) — susceptible fraction",
           hjust = 0, colour = "#d62728", size = 5.5, fontface = "bold") +
  annotate("curve", x = 4.4, y = 0.45, xend = 3.8, yend = 0.39,
           arrow = arrow(length = unit(0.18, "cm")),
           curvature = -0.2, colour = "#d62728") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "time", y = NULL,
       title = "One curve, two names")
```
:::
::::

## How is it different?

Five twists. **One per section.**

::: {.incremental}
1. **The hazard is endogenous.** It depends on $I(t)$.   ◂ § 1
2. **Compartments couple individuals.** Multistate with feedback.   ◂ § 2
3. **The population sets the clock.** $R_0$, $R_t$, final size.   ◂ § 3
4. **We see only fragments.** Censored, truncated, delayed data.   ◂ § 4
5. **Interventions act on populations.** Serology, vaccination, mortality.   ◂ § 5
:::

::: {.fragment .fade-in}
> **Same math, applied to a population with feedback.**
:::

::: {.notes}
Each twist gets one section. By the end, the audience should be able to tell you which twist they're inside.
:::

## Roadmap

```{mermaid}
%%| fig-width: 10
flowchart LR
  B1["<b>§ 1 — Twist 1</b><br/>endogenous hazard<br/><i>force of infection</i>"]:::bridge
  B2["<b>§ 2 — Twist 2</b><br/>coupled compartments<br/><i>SIR / SEIR</i>"]:::bridge
  B3["<b>§ 3 — Twist 3</b><br/>population clock<br/><i>R<sub>0</sub>, R<sub>t</sub>, renewal</i>"]:::bridge
  B4["<b>§ 4 — Twist 4</b><br/>fragmented data<br/><i>censoring &amp; truncation</i>"]:::bridge
  B5["<b>§ 5 — Twist 5</b><br/>population interventions<br/><i>case-study toolkit</i>"]:::bridge
  B1 --> B2 --> B3 --> B4 --> B5
  classDef bridge fill:#fff2cc,stroke:#b58900,color:#5a4500
```

# Twist 1 — The hazard is endogenous {background-color="#1f77b4" style="color:white;"}

::: {style="color:white;"}
**§ 1.** Force of infection — the hazard $\lambda(t)$ that depends on what everyone else is doing.
:::

## The hazard, refreshed

:::: {.columns}
::: {.column width="55%"}
$$h(t) = \lim_{\Delta t \downarrow 0} \frac{\Pr(t \le T < t+\Delta t \mid T \ge t)}{\Delta t}$$

$$S(t) = \exp\!\left\{ -\!\int_0^t h(u)\, du \right\}$$

A single object — the hazard — generates the survival curve, the density, the cumulative hazard.
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
t <- seq(0, 8, by = 0.02)
h <- 0.15 + 0.06 * t
H <- cumsum(h) * 0.02
S <- exp(-H)
df <- tibble(t, hazard = h, survival = S) |>
  pivot_longer(-t)
ggplot(df, aes(t, value)) +
  geom_line(linewidth = 1.4, colour = "#1f77b4") +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(x = "time", y = NULL)
```
:::
::::

## The force of infection

:::: {.columns}
::: {.column width="55%"}
Let $S(t)$ = fraction of the population still susceptible.

$$\lambda(t) = -\frac{1}{S(t)} \frac{dS(t)}{dt}$$

$$S(t) = \exp\!\left\{ -\!\int_0^t \lambda(u)\, du \right\}$$

Identical to the hazard — except the "subjects" are people who have not yet been infected, and $\lambda(t)$ is *endogenous*.

::: {.callout-note appearance="minimal"}
Under **homogeneous mixing**, the population FoI equals the per-susceptible hazard. Heterogeneity breaks this — see the frailty slide.
:::
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
# constant force of infection lambda = 0.25
t <- seq(0, 12, 0.05)
S <- exp(-0.25 * t)
lambda <- rep(0.25, length(t))
df <- tibble(t, susceptible = S, lambda = lambda) |>
  pivot_longer(-t)
ggplot(df, aes(t, value)) +
  geom_line(linewidth = 1.4, colour = "#d62728") +
  facet_wrap(~name, ncol = 1, scales = "free_y",
             labeller = as_labeller(c(susceptible = "S(t)  susceptible fraction",
                                      lambda = "lambda(t)  force of infection"))) +
  labs(x = "time", y = NULL)
```
:::
::::

## A first historic anchor: Bernoulli, 1760

:::: {.columns}
::: {.column width="55%"}
Daniel Bernoulli wrote down a survival model for smallpox **two centuries before Cox**.

::: {.incremental}
- Two competing hazards
- $\lambda$ : *catching* smallpox
- $\mu(a)$ : dying from anything else at age $a$
- Life expectancy *with* and *without* smallpox
- A competing-risks survival model — first one ever written
:::
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
a <- seq(0, 80, 0.5)
lambda_sp <- 0.0125
mu       <- 0.003 + 0.0008 * a^1.05
S_obs   <- exp(-cumsum(lambda_sp + mu) * 0.5)
S_no_sp <- exp(-cumsum(mu) * 0.5)
gain <- round(sum(S_no_sp - S_obs) * 0.5, 1)  # area = years gained
df <- tibble(a, observed = S_obs, eliminated = S_no_sp)
ggplot(df, aes(a)) +
  geom_ribbon(aes(ymin = observed, ymax = eliminated),
              fill = "#fff2cc", alpha = 0.9) +
  geom_line(aes(y = observed),    colour = "#d62728", linewidth = 1.5) +
  geom_line(aes(y = eliminated),  colour = "#1f77b4", linewidth = 1.5) +
  annotate("text", x = 55, y = 0.55,
           label = paste0("life-years gained\nper person ≈ ", gain),
           colour = "#5a4500", size = 4.5, fontface = "bold") +
  annotate("text", x = 18, y = 0.97, label = "if smallpox eliminated",
           hjust = 0, colour = "#1f77b4", size = 4.5, fontface = "bold") +
  annotate("text", x = 18, y = 0.78, label = "observed",
           hjust = 0, colour = "#d62728", size = 4.5, fontface = "bold") +
  labs(x = "age (years)", y = "S(a)",
       title = "Bernoulli's counterfactual, redrawn")
```
:::
::::

## Heterogeneity = frailty {.smaller}

:::: {.columns}
::: {.column width="50%"}
**Cox model:**
$$h_i(t) = h_0(t)\, e^{\beta'x_i}$$

**Mixed-effects / frailty:**
$$h_i(t) = h_0(t)\, Z_i,\quad Z_i \sim \pi(z)$$

**Heterogeneous FoI:**
$$\lambda_i(t) = c_i\, \beta\, \frac{I(t)}{N}$$

The $c_i$ play the role of frailties.
:::
::: {.column width="50%"}
```{r}
#| fig-height: 4.7
set.seed(1)
t <- seq(0, 12, 0.05)
# population FoI = 0.25, but 3 sub-groups with multipliers
groups <- tibble(g = c("low (c=0.4)", "avg (c=1)", "high (c=2.5)"),
                 c = c(0.4, 1, 2.5))
df <- expand.grid(t = t, g = groups$g) |>
  left_join(groups, by = "g") |>
  mutate(S = exp(-0.25 * c * t))
pop_avg <- df |> group_by(t) |>
  summarise(S = mean(S), .groups = "drop") |> mutate(g = "population mean")
ggplot() +
  geom_line(data = df, aes(t, S, colour = g), linewidth = 1.2) +
  geom_line(data = pop_avg, aes(t, S), colour = "black", linewidth = 1.6, linetype = "22") +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "time", y = "S(t)", colour = NULL,
       title = "High-frailty subjects deplete first",
       subtitle = "Black: equal-weight mixture. Apparent FoI declines without any intervention.") +
  theme(legend.position = "bottom")
```
:::
::::

::: {.fragment}
The "selection of the susceptibles" is the same artefact as frailty-induced declining hazards in survival data.
:::

# Twist 2 — Compartments couple individuals {background-color="#2ca02c" style="color:white;"}

::: {style="color:white;"}
**§ 2.** Multistate models with feedback — SIR, SEIR, competing risks.
:::

## SIR is a multistate model

```{mermaid}
%%| fig-width: 9
flowchart LR
  S((S)) -->|"&lambda;(t) = &beta; I/N"| I((I))
  I -->|"&gamma;"| R((R))
  classDef s fill:#cfe2f3,stroke:#1f77b4,stroke-width:2px
  classDef i fill:#fde2e2,stroke:#d62728,stroke-width:2px
  classDef r fill:#d9f1d9,stroke:#2ca02c,stroke-width:2px
  class S s
  class I i
  class R r
```

:::: {.columns}
::: {.column width="50%"}
**Multistate-model view**

- States: $\{S, I, R\}$
- Transition intensities: $q_{SI}(t),\ q_{IR}$
- Aalen / Andersen–Gill machinery applies *— with the twist on the next slide*
:::
::: {.column width="50%"}
**Epidemiologist's view**

- Compartments
- Rate equations
- "Mass action" coupling: $q_{SI}(t)$ depends on $I(t)$
:::
::::

## SIR as ODEs (Kermack–McKendrick, 1927)

:::: {.columns}
::: {.column width="48%"}
$$
\begin{aligned}
\frac{dS}{dt} &= -\beta\, \tfrac{SI}{N} \\
\frac{dI}{dt} &= +\beta\, \tfrac{SI}{N} - \gamma I \\
\frac{dR}{dt} &= +\gamma I
\end{aligned}
$$

The same multistate diagram, written as a system of ODEs.

A landmark 1927 paper that shaped a century of epidemiology.
:::
::: {.column width="52%"}
```{r}
#| fig-height: 4.7
sir <- function(t, y, p) {
  with(as.list(c(y, p)), {
    dS <- -beta * S * I / N
    dI <-  beta * S * I / N - gamma * I
    dR <-  gamma * I
    list(c(dS, dI, dR))
  })
}
N <- 1
y0 <- c(S = 0.999, I = 0.001, R = 0)
out <- as.data.frame(ode(y0, seq(0, 120, 0.5), sir,
                         parms = c(beta = 0.4, gamma = 0.15, N = N)))
out |>
  pivot_longer(-time, names_to = "state") |>
  ggplot(aes(time, value, colour = state)) +
  geom_line(linewidth = 1.6) +
  scale_colour_manual(values = pal) +
  labs(x = "day", y = "fraction of population",
       title = "SIR epidemic curve",
       subtitle = "beta = 0.4, gamma = 0.15, R0 = 2.67",
       colour = NULL) +
  theme(legend.position = "bottom")
```
:::
::::

## What makes it different from survival? {.smaller}

:::: {.columns}
::: {.column width="50%"}
**Survival analysis**

$$h(t) \text{ is exogenous}$$

The risk set shrinks because subjects experience the event.
:::
::: {.column width="50%"}
**Epidemic models**

$$\lambda(t) = \beta\, \frac{I(t)}{N}\ \text{is endogenous}$$

The hazard rises and then falls because $I(t)$ does — a feedback loop.
:::
::::

```{mermaid}
%%| fig-width: 9
flowchart LR
  X1["I(t)"] -->|"raises"| L1["&lambda;(t)"]
  L1 -->|"depletes"| S1["S(t)"]
  S1 -->|"feeds new I"| X1
  classDef k fill:#fff2cc,stroke:#b58900,stroke-width:2px
  class X1,L1,S1 k
```

::: {.fragment .fade-in}
**Why naive Cox regression on epidemic data is dangerous:**
individuals' hazards are *correlated* through shared $I(t)$. Independence breaks, effective sample size shrinks, and covariate effects can be confounded with epidemic timing.
:::

## Sojourn times = waiting-time distributions

:::: {.columns}
::: {.column width="50%"}
The classic SIR assumes
$$\text{Time in I} \sim \text{Exp}(\gamma).$$

Constant transition intensity ⇒ memoryless ⇒ **Markov**.

Real infectious periods are usually **gamma-shaped**.

The fix: chain $k$ exponential sub-stages — the "method of stages".
This is exactly a phase-type distribution / semi-Markov multistate model.
:::
::: {.column width="50%"}
```{r}
#| fig-height: 4.7
t <- seq(0, 20, 0.05)
df <- bind_rows(
  tibble(t, density = dexp(t, rate = 1/6),     dist = "Exponential (k=1)"),
  tibble(t, density = dgamma(t, shape = 3, rate = 3/6), dist = "Gamma (k=3)"),
  tibble(t, density = dgamma(t, shape = 6, rate = 6/6), dist = "Gamma (k=6)")
)
ggplot(df, aes(t, density, colour = dist)) +
  geom_line(linewidth = 1.4) +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "days infectious", y = "density", colour = NULL,
       title = "Same mean, very different shapes") +
  theme(legend.position = "bottom")
```
:::
::::

## SEIR: an extra latent state {.smaller}

```{mermaid}
%%| fig-width: 10
flowchart LR
  S((S)) -->|"&lambda;(t)"| E((E))
  E -->|"&sigma;"| I((I))
  I -->|"&gamma;"| R((R))
  classDef s fill:#cfe2f3,stroke:#1f77b4
  classDef e fill:#e8d8f7,stroke:#9467bd
  classDef i fill:#fde2e2,stroke:#d62728
  classDef r fill:#d9f1d9,stroke:#2ca02c
  class S s
  class E e
  class I i
  class R r
```

:::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- **E** = exposed but not yet infectious (latency)
- Sojourn $E \to I$ : incubation period
- Sojourn $I \to R$ : infectious period
- Both are survival distributions you would estimate from line-list data
:::
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
seir <- function(t, y, p) {
  with(as.list(c(y, p)), {
    dS <- -beta * S * I / N
    dE <-  beta * S * I / N - sigma * E
    dI <-  sigma * E - gamma * I
    dR <-  gamma * I
    list(c(dS, dE, dI, dR))
  })
}
N <- 1
y0 <- c(S = 0.999, E = 0, I = 0.001, R = 0)
out <- as.data.frame(ode(y0, seq(0, 150, 0.5), seir,
                         parms = c(beta = 0.5, sigma = 1/4, gamma = 1/7, N = N)))
out |>
  pivot_longer(-time, names_to = "state") |>
  ggplot(aes(time, value, colour = state)) +
  geom_line(linewidth = 1.6) +
  scale_colour_manual(values = pal) +
  labs(x = "day", y = "fraction", colour = NULL,
       title = "SEIR dynamics") +
  theme(legend.position = "bottom")
```
:::
::::

## Competing risks: I → R vs I → D

```{mermaid}
%%| fig-width: 9
flowchart LR
  S((S)) -->|"&lambda;(t)"| I((I))
  I -->|"&gamma;(1-IFR)"| R((R))
  I -->|"&gamma; * IFR"| D((D))
  classDef s fill:#cfe2f3,stroke:#1f77b4
  classDef i fill:#fde2e2,stroke:#d62728
  classDef r fill:#d9f1d9,stroke:#2ca02c
  classDef d fill:#dddddd,stroke:#444
  class S s
  class I i
  class R r
  class D d
```

::: {.fragment}
- **Infection fatality ratio (IFR)** is just the cause-specific cumulative incidence at $\infty$.
- Same identifiability traps as in cause-specific survival: do not confuse cause-specific hazards with subdistribution hazards.
:::

## A discrete-time cousin: Reed–Frost (1928) {.smaller}

:::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- Time advances in *generations* (not continuously)
- $I_{t+1} \mid S_t, I_t \sim \text{Bin}\!\left(S_t,\ 1 - (1-p)^{I_t}\right)$
- Each susceptible avoids infection if it escapes contact with all $I_t$ infectives
- $1 - (1-p)^{I_t}$ is the *discrete-time analogue of* $1 - e^{-\Lambda_t}$
- Discrete survival $\Leftrightarrow$ chain-binomial epidemic
:::

::: {.fragment .fade-in}
**Stochasticity matters most when $I_t$ is small** — early or late in an outbreak — the same regime where survival cares about ties and small risk sets.
:::
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
set.seed(7)
N <- 1000; p <- 0.0025
sim <- function() {
  S <- N - 1; I <- 1
  out <- tibble(gen = 0, S = S, I = I)
  for (g in 1:25) {
    new <- rbinom(1, S, 1 - (1 - p)^I)
    S <- S - new; I <- new
    out <- bind_rows(out, tibble(gen = g, S = S, I = I))
    if (I == 0) break
  }
  out
}
runs <- bind_rows(lapply(1:30, function(i) sim() |> mutate(rep = i)))
ggplot(runs, aes(gen, I, group = rep)) +
  geom_line(alpha = 0.35, colour = "#d62728") +
  labs(x = "generation", y = "new infections",
       title = "30 stochastic Reed-Frost runs")
```
:::
::::

# Twist 3 — The population sets the clock {background-color="#9467bd" style="color:white;"}

::: {style="color:white;"}
**§ 3.** $R_0$, $R_t$, generation interval, renewal equation, final size.
:::

## R₀ — the branching factor

:::: {.columns}
::: {.column width="55%"}
$$R_0 = \text{expected secondary cases from one infective in a fully susceptible population.}$$

For the simple SIR:
$$R_0 = \frac{\beta}{\gamma} = \underbrace{c}_{\text{contacts}} \cdot \underbrace{p}_{\text{transmission prob.}} \cdot \underbrace{D}_{\text{mean duration}}$$

Each factor is something a survival analyst recognises — especially $D$, a mean sojourn time.
:::
::: {.column width="45%"}
```{mermaid}
%%| fig-width: 6
flowchart TB
  A((1)) --> B1((2))
  A --> B2((2))
  A --> B3((2))
  B1 --> C1((3))
  B1 --> C2((3))
  B1 --> C3((3))
  B2 --> C4((3))
  B2 --> C5((3))
  B2 --> C6((3))
  B3 --> C7((3))
  B3 --> C8((3))
  B3 --> C9((3))
```

A geometric tree with branching $R_0$.
:::
::::

## R₀ as an integral over the infectious period

$$
R_0 = \int_0^\infty \beta(\tau)\, S_I(\tau)\, d\tau
$$

:::: {.columns}
::: {.column width="55%"}
- $S_I(\tau)$ = probability still infectious $\tau$ after onset = a **survival function**
- $\beta(\tau)$ = age-of-infection contact/transmission rate
- $R_0$ = area under "infectiousness profile $\times$ survival of infectiousness"

When $\beta$ is constant, this collapses to
$$R_0 = \beta \cdot E[T_{\text{infectious}}]$$
— the contact-transmission rate times the **expected infectious duration**, exactly what you'd write in a survival analysis.
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
tau <- seq(0, 14, 0.05)
S_I <- 1 - pgamma(tau, shape = 6, rate = 6/7)   # mean ~7d
beta_tau <- 0.35 * dgamma(tau, shape = 4, rate = 1.2) /
  max(dgamma(tau, shape = 4, rate = 1.2))
df <- tibble(tau, infectiousness = beta_tau, survival = S_I,
             product = beta_tau * S_I) |>
  pivot_longer(-tau)
ggplot(df, aes(tau, value, colour = name)) +
  geom_line(linewidth = 1.4) +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "time since infection (days)", y = NULL, colour = NULL,
       title = "R0 = integral of beta(tau) * S_I(tau)") +
  theme(legend.position = "bottom")
```
:::
::::

## Effective reproduction number Rₜ

:::: {.columns}
::: {.column width="55%"}
$$R_t = R_0 \cdot \frac{S(t)}{N}$$

::: {.incremental}
- Susceptibles deplete ⇒ $R_t$ falls
- Interventions reduce $\beta$ ⇒ $R_t$ falls
- Outbreak ends when $R_t < 1$
:::

Wallinga–Lipsitch identity (early phase):
$$\frac{1}{R} = \int_0^\infty g(\tau)\, e^{-r\tau}\,d\tau$$

where $r$ is the exponential growth rate and $g(\tau)$ the generation interval — the moment generating function of a survival distribution.
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
out <- as.data.frame(ode(c(S = 0.999, I = 0.001, R = 0),
                         seq(0, 120, 0.5), sir,
                         parms = c(beta = 0.4, gamma = 0.15, N = 1)))
out$Rt <- (0.4/0.15) * out$S
ggplot(out, aes(time, Rt)) +
  geom_hline(yintercept = 1, linetype = "22") +
  geom_line(linewidth = 1.6, colour = "#9467bd") +
  labs(x = "day", y = expression(R[t]),
       title = "Rt declines with susceptible depletion")
```
:::
::::

## Final-size relation & herd immunity {.smaller}

:::: {.columns}
::: {.column width="55%"}
For the simple SIR, the fraction $s_\infty$ that *escapes* infection satisfies the self-consistency equation
$$s_\infty = e^{-R_0(1 - s_\infty)}.$$

A pure survival statement: it is a survival probability, expressed as the exponential of a cumulative force of infection.

**Herd immunity threshold:**
$$1 - \tfrac{1}{R_0}$$
— the depleted-susceptible fraction at which $R_t$ first crosses 1.
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
R0_grid <- seq(1.01, 6, 0.02)
final_size <- function(R0) {
  f <- function(s) s - exp(-R0 * (1 - s))
  uniroot(f, c(1e-8, 1 - 1e-8))$root
}
df <- tibble(
  R0 = R0_grid,
  `attack rate (1 - s_inf)` = 1 - sapply(R0_grid, final_size),
  `herd immunity threshold` = 1 - 1 / R0_grid
) |> pivot_longer(-R0)
ggplot(df, aes(R0, value, colour = name)) +
  geom_line(linewidth = 1.6) +
  scale_colour_manual(values = c(`attack rate (1 - s_inf)` = "#d62728",
                                 `herd immunity threshold` = "#1f77b4")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(x = expression(R[0]), y = NULL, colour = NULL,
       title = "Two consequences of R0",
       subtitle = "Attack rate (red) overshoots the herd-immunity threshold (blue)") +
  theme(legend.position = "bottom")
```
:::
::::

## The renewal equation

$$
i(t) \;=\; \int_0^\infty i(t-\tau)\, R_{t-\tau}\, g(\tau)\, d\tau
$$

:::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- $i(t)$ : incidence at time $t$
- $g(\tau)$ : **generation interval** density
- $R_{t-\tau}$ : case reproduction number (Wallinga–Teunis)
:::

This is a **convolution** — and a deconvolution problem when $R$ is unknown.

It is also the integral form of the Kermack–McKendrick model — predating the ODE system in the same 1927 paper.

::: {.callout-note appearance="minimal"}
Two conventions: **case** $R_{t-\tau}$ (looks back in time, Wallinga–Teunis) and **instantaneous** $R_t$ (looks forward, Cori et al.). They coincide in steady state.
:::
:::
::: {.column width="45%"}
```{mermaid}
%%| fig-width: 6
flowchart LR
  P["past incidence i(t-&tau;)"] -->|" * "| G["generation interval g(&tau;)"]
  G --> CV["convolve"]
  CV --> N["new incidence i(t)"]
  R["R<sub>t</sub>"] --> CV
  classDef k fill:#fff2cc,stroke:#b58900
  class P,G,R,CV,N k
```
:::
::::

## The generation interval is a survival distribution

:::: {.columns}
::: {.column width="55%"}
The generation interval $\tau$ is the time between infection of the **infector** and infection of the **infectee**.

Conditional on a contact event, it inherits its shape from:

- the time-to-infection of the contact (a hazard) and
- the infector's infectiousness profile.

You estimate it the way you would estimate any time-to-event distribution — from doubly-censored, biased samples.
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
tau <- seq(0, 14, 0.05)
df <- tibble(
  tau,
  density = dgamma(tau, shape = 4, rate = 4/5.2)
)
ggplot(df, aes(tau, density)) +
  geom_area(fill = "#f1c1c1", colour = "#d62728", linewidth = 1.3) +
  labs(x = "days between infections", y = "density",
       title = "Generation interval g(tau)",
       subtitle = "Gamma-shaped, mean ~5 days (illustrative)")
```
:::
::::

## Serial interval vs generation interval {.smaller}

:::: {.columns}
::: {.column width="55%"}
- **Generation interval (GI):** time between *infection of infector* and *infection of infectee*. Governs the renewal equation.
- **Serial interval (SI):** time between *symptom onset of infector* and *symptom onset of infectee*. Observable in line lists.

$$\text{SI} = \text{GI} + (\text{incubation}_{\text{infectee}} - \text{incubation}_{\text{infector}})$$

Equal means, but SI has **larger variance** and can be negative.

::: {.callout-note appearance="minimal"}
In the workshop case study, the contact-tracing data gives both — onset times always, infection times only sometimes (interval-censored).
:::
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
set.seed(9)
n <- 5000
inc_or <- rgamma(n, shape = 4, rate = 4/4)   # infector incubation
inc_ee <- rgamma(n, shape = 4, rate = 4/4)   # infectee incubation
GI <- rgamma(n, shape = 4, rate = 4/5.2)
SI <- GI + (inc_ee - inc_or)
df <- bind_rows(
  tibble(value = GI, kind = "Generation interval"),
  tibble(value = SI, kind = "Serial interval")
)
ggplot(df, aes(value, fill = kind, colour = kind)) +
  geom_density(alpha = 0.35, linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "22") +
  scale_fill_manual(values  = c("Generation interval" = "#d62728",
                                "Serial interval"    = "#1f77b4")) +
  scale_colour_manual(values = c("Generation interval" = "#d62728",
                                 "Serial interval"    = "#1f77b4")) +
  labs(x = "days", y = "density", fill = NULL, colour = NULL,
       title = "Same mean, fatter tails — and SI can be negative") +
  theme(legend.position = "bottom")
```
:::
::::

## Estimating R₀ from the early-phase growth {.smaller}

:::: {.columns}
::: {.column width="55%"}
Early in the outbreak, incidence grows as $i(t) \approx i_0 e^{rt}$.

**Euler–Lotka** turns the growth rate $r$ into $R_0$ via the generation interval:
$$\frac{1}{R_0} = \int_0^\infty e^{-r\tau}\, w(\tau)\, d\tau$$

For a Gamma generation time $w(\tau)\sim \text{Gamma}(\alpha,\beta)$, the Laplace transform gives the closed form
$$\widehat{R}_0 = (1 + \widehat{\beta}\, r)^{\widehat{\alpha}}.$$

A two-step recipe: fit GI from contact tracing → estimate $r$ from early incidence → plug in.
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
r_grid <- seq(0, 0.35, 0.005)
# Gamma GI: mean = 5.2, sd = 2.6 → shape=4, scale=1.3
alpha <- 4; beta <- 1.3
R0 <- (1 + beta * r_grid)^alpha
df <- tibble(r = r_grid, R0 = R0)
ggplot(df, aes(r, R0)) +
  geom_line(linewidth = 1.6, colour = "#9467bd") +
  geom_hline(yintercept = 1, linetype = "22") +
  annotate("text", x = 0.27, y = 8,
           label = "R0 = (1 + beta * r)^alpha\nGamma GI, mean 5.2 d",
           hjust = 1, size = 4.2, fontface = "italic") +
  labs(x = "exponential growth rate r (per day)", y = expression(widehat(R)[0]),
       title = "Growth rate to R0, via the GI Laplace transform")
```
:::
::::

## A historic Rₜ: 1918 influenza, Philadelphia {.smaller}

:::: {.columns}
::: {.column width="48%"}
- Modern reanalyses (Bootsma & Ferguson 2007, Hatchett et al. 2007) apply the renewal equation to city-level death counts
- $R_t$ rose above 2 in early October 1918, then fell rapidly under NPIs **and** depletion
- The Philadelphia–St-Louis comparison is the canonical "early NPIs flatten the curve" story
:::
::: {.column width="52%"}
```{r}
#| fig-height: 5.0
set.seed(2)
days <- 0:60
peak <- 28
mu <- 350 * dnorm(days, peak, 9) / max(dnorm(days, peak, 9))
deaths <- rpois(length(days), pmax(mu, 0.5))
# Stylised Rt: rises above 2, drops rapidly to ~0.7 around peak
Rt <- 2.4 / (1 + exp(0.25 * (days - 22)))   # smooth fall
Rt <- pmax(Rt, 0.6)
df_d <- tibble(day = days, value = deaths, panel = "weekly deaths")
df_r <- tibble(day = days, value = Rt,     panel = "R_t")
df <- bind_rows(df_d, df_r) |>
  mutate(panel = factor(panel, levels = c("weekly deaths", "R_t")))
ggplot(df, aes(day, value)) +
  geom_col(data = subset(df, panel == "weekly deaths"),
           fill = "#d62728", alpha = 0.85) +
  geom_hline(data = subset(df, panel == "R_t"),
             aes(yintercept = 1), linetype = "22", colour = "grey40") +
  geom_line(data = subset(df, panel == "R_t"),
            colour = "#9467bd", linewidth = 1.6) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  labs(x = "days since first death (illustrative)", y = NULL,
       title = "Stylised 1918 Philadelphia wave")
```
:::
::::

# Twist 4 — We see only fragments {background-color="#d62728" style="color:white;"}

::: {style="color:white;"}
**§ 4.** Censored exposures, truncated incidence, reporting delays.
:::

## A second historic anchor: John Snow, 1854 {.smaller}

:::: {.columns}
::: {.column width="55%"}
Cholera outbreak, Broad Street, Soho. Snow:

- Reconstructed **exposure times** from interviews
- Plotted cases against a *risk set* (households per water source)
- Inferred a hazard ratio between two water companies of ≈ **8.5**

A door-to-door **cohort study with interval-censored exposure** — done in 1854 with no statistical theory yet written.

::: {.fragment}
> The first epidemiological estimator of a hazard ratio was a survival analyst doing nowcasting under right-truncation, by hand.
:::
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
df <- tibble(
  company = c("Southwark & Vauxhall\n(contaminated)",
              "Lambeth\n(clean source)"),
  houses  = c(40046, 26107),
  deaths  = c(1263, 98),
  rate    = deaths / houses * 10000
)
ggplot(df, aes(reorder(company, rate), rate, fill = company)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.0f / 10k", rate)),
            hjust = -0.1, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("#1f77b4", "#d62728")) +
  coord_flip(clip = "off") +
  expand_limits(y = max(df$rate) * 1.25) +
  labs(x = NULL, y = "cholera deaths per 10,000 houses",
       title = "Snow's natural experiment, 1854") +
  theme(plot.margin = margin(5, 60, 5, 5))
```
:::
::::

## The line list {.smaller}

```{mermaid}
%%| fig-width: 10
flowchart LR
  Inf["infection<br/>(unobserved)"] --> Sym["symptom onset"]
  Sym --> Conf["confirmation /<br/>report date"]
  Conf --> End["outcome<br/>(R or D)"]
  classDef u fill:#eeeeee,stroke:#888,stroke-dasharray:4 4
  classDef o fill:#d9f1d9,stroke:#2ca02c
  class Inf u
  class Sym,Conf,End o
```

::: {.fragment}
::: {.incremental}
- The infection time is almost **never** observed
- The line list gives you onset, confirmation, outcome
- Every duration of interest (incubation, generation, onset-to-death) involves at least one **censored** endpoint
:::
:::

## Doubly interval-censored: incubation period

:::: {.columns}
::: {.column width="55%"}
For each case we know an exposure window $[E_L, E_R]$ and a symptom-onset window $[O_L, O_R]$.

The incubation period is
$$T = O - E,\qquad E\in[E_L,E_R],\ O\in[O_L,O_R].$$

This is a textbook **doubly interval-censored** survival problem.
EM, Bayesian augmentation, multiple imputation — the standard kit.
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
set.seed(11)
n <- 12
df <- tibble(
  case = factor(seq_len(n)),
  EL = runif(n, -3, -1), ER = EL + runif(n, 0.5, 2),
  OL = ER + runif(n, 1.5, 6), OR = OL + runif(n, 0.3, 1.5)
)
ggplot(df) +
  geom_segment(aes(x = EL, xend = ER, y = case, yend = case),
               colour = "#1f77b4", linewidth = 3, lineend = "round") +
  geom_segment(aes(x = OL, xend = OR, y = case, yend = case),
               colour = "#d62728", linewidth = 3, lineend = "round") +
  labs(x = "days", y = "case",
       title = "Exposure (blue) and onset (red) intervals",
       subtitle = "Both endpoints censored")
```
:::
::::

## Right-truncation in real time

:::: {.columns}
::: {.column width="50%"}
**Survival mindset:** a subject's event happened, we just stopped following.

**Outbreak mindset:** a case happened, but **hasn't been reported yet** because of reporting delay $D$.

If we condition on cases reported by today, we systematically miss those with long delays — **right-truncation**.

This is the classic HIV back-calculation problem (Brookmeyer & Gail, 1980s).
:::
::: {.column width="50%"}
```{r}
#| fig-height: 4.7
set.seed(3)
days <- 0:80
mu <- 160 * dnorm(days, 50, 9) / max(dnorm(days, 50, 9))
true_inc <- rpois(length(days), pmax(mu, 0.4))
delay_pmf <- dgamma(0:30, shape = 2, rate = 0.4)
delay_pmf <- delay_pmf / sum(delay_pmf)
today <- 70
reported <- rep(0, length(days))
for (d in seq_along(days)) {
  for (k in 0:30) {
    if (days[d] + k <= today)
      reported[d] <- reported[d] + true_inc[d] * delay_pmf[k+1]
  }
}
df <- tibble(day = days, true = true_inc, reported = reported)
ggplot(df, aes(day)) +
  geom_col(aes(y = true), fill = NA, colour = "#1f77b4", linewidth = 0.4) +
  geom_col(aes(y = reported), fill = "#d62728", alpha = 0.85) +
  geom_vline(xintercept = today, linetype = "22") +
  annotate("text", x = today + 0.5, y = max(true_inc) * 0.95,
           label = "← today", hjust = 0, size = 5, fontface = "bold") +
  annotate("text", x = 8, y = max(true_inc) * 0.6,
           label = "true (outline)", colour = "#1f77b4",
           hjust = 0, size = 4.5, fontface = "bold") +
  annotate("text", x = 8, y = max(true_inc) * 0.45,
           label = "what we see today (filled)", colour = "#d62728",
           hjust = 0, size = 4.5, fontface = "bold") +
  labs(x = "day", y = "cases",
       title = "Right-truncated incidence")
```
:::
::::

## The Rₜ wobble at the right edge {.smaller}

:::: {.columns}
::: {.column width="50%"}
A naive $R_t$ estimator that ignores reporting delays will produce a sharp **artefactual drop** at the most recent dates.

This is identical to estimating a hazard at the right edge of follow-up without accounting for censoring.

Solutions = **nowcasting**:

::: {.incremental}
- delay-distribution deconvolution
- Bayesian state-space models (`EpiNow2`, `epinowcast`)
- back-calculation
:::
:::
::: {.column width="50%"}
```{r}
#| fig-height: 4.7
set.seed(4)
days <- 0:60
true_Rt <- 1.2 + 0.4 * sin(days / 8)
naive_Rt <- true_Rt
naive_Rt[55:61] <- naive_Rt[55:61] - seq(0.05, 0.7, length.out = 7)
df <- tibble(day = days, true = true_Rt, `naive (right-truncated)` = naive_Rt) |>
  pivot_longer(-day)
ggplot(df, aes(day, value, colour = name)) +
  geom_hline(yintercept = 1, linetype = "22") +
  geom_line(linewidth = 1.5) +
  scale_colour_manual(values = c(true = "#1f77b4",
                                 `naive (right-truncated)` = "#d62728")) +
  labs(x = "day", y = expression(R[t]), colour = NULL,
       title = "Why the latest Rt looks too low") +
  theme(legend.position = "bottom")
```
:::
::::

## Nowcasting in one slide

```{mermaid}
%%| fig-width: 10
flowchart LR
  R["reported<br/>cases by date"] --> M["delay<br/>model f(d)"]
  M --> D["deconvolution"]
  D --> N["nowcast of<br/>true incidence"]
  N --> Rt["unbiased R<sub>t</sub>"]
  classDef k fill:#fff2cc,stroke:#b58900
  class R,M,D,N,Rt k
```

::: {.fragment}
> Nowcasting is to outbreak surveillance what censoring corrections are to follow-up studies.
:::

## Other censoring/truncation cousins {.smaller}

| Outbreak situation | Survival analogue |
|---|---|
| Reporting delay | Right-truncation |
| Cases never confirmed | Right-censoring at study end |
| Exposure window unknown | Left-truncation / interval censoring |
| Doubly-censored incubation | Doubly interval-censored data |
| Onset-to-death cohort still alive | Right-censored time-to-event |
| Asymptomatic infections | Latent / unobserved events |

::: {.fragment}
Every method on the right has been used to fix the corresponding problem on the left, often by your own community.
:::

# Twist 5 — Interventions act on populations {background-color="#e08214" style="color:white;"}

::: {style="color:white;"}
**§ 5.** Serology, vaccine effectiveness, excess mortality — the case-study toolkit.
:::

## Serology: a window on the cumulative hazard {.smaller}

:::: {.columns}
::: {.column width="55%"}
A repeated cross-sectional serosurvey gives **time-dependent seroprevalence** $\pi_s(t) = \Pr(Z=1 \mid T_s = t)$.

For a perfect test, $\pi_s(t)$ is the **cumulative incidence** by time $t$ — i.e. $1 - S(t)$ where $S$ is the susceptible-survival function.

**Imperfect test (Rogan–Gladen, 1978):**
$$\pi(t) = \frac{\pi_s(t) + \psi - 1}{\phi + \psi - 1}$$

with sensitivity $\phi$, specificity $\psi$ — a textbook misclassification correction.

::: {.fragment}
*Case-study task 4:* combine serial seroprevalence with reported incidence to estimate **under-reporting**.
:::
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
# Stylised serosurvey: 4 sampling times, R0 ~ 2, single SIR wave
out <- as.data.frame(ode(c(S = 0.999, I = 0.001, R = 0),
                         seq(0, 80, 0.5), sir,
                         parms = c(beta = 0.45, gamma = 0.18, N = 1)))
out$cum_inc <- 1 - out$S
sample_t <- c(10, 20, 30, 40)
set.seed(5)
samp <- tibble(t = sample_t,
               true = approx(out$time, out$cum_inc, sample_t)$y) |>
  mutate(observed = pmin(1, pmax(0, true + rnorm(n(), 0, 0.015))))
ggplot() +
  geom_line(data = out, aes(time, cum_inc),
            colour = "#1f77b4", linewidth = 1.4) +
  geom_point(data = samp, aes(t, observed),
             size = 5, colour = "#d62728") +
  geom_errorbar(data = samp,
                aes(x = t, ymin = pmax(0, observed - 0.03),
                    ymax = pmin(1, observed + 0.03)),
                width = 1, colour = "#d62728", linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "day", y = "cumulative incidence / seroprevalence",
       title = "Cross-sectional serosurveys",
       subtitle = "Blue: true 1 - S(t). Red: noisy survey estimates.")
```
:::
::::

## Mixture models on antibody titers {.smaller}

:::: {.columns}
::: {.column width="55%"}
Skip the threshold. Model the **continuous** log-titer $Y_i$ as a two-component mixture:
$$f(y \mid t) = (1-\pi(t))\, f_{\text{neg}}(y) + \pi(t)\, f_{\text{pos}}(y)$$

- $f_{\text{neg}}, f_{\text{pos}}$ : titre distributions in seronegative / seropositive subpopulations
- $\pi(t)$ : true prevalence — the survival quantity you want
- **Left-censoring at the limit of detection** ($\ln 5 \approx 1.61$) — standard survival-analysis treatment

A nonparametric or parametric mixture cleanly propagates uncertainty about sensitivity / specificity into the seroprevalence estimate.
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
set.seed(13)
y <- seq(0, 8, 0.02)
f_neg <- dnorm(y, mean = 1.8, sd = 0.5)
f_pos <- dnorm(y, mean = 5.5, sd = 0.9)
pi_t <- 0.35
mix <- (1 - pi_t) * f_neg + pi_t * f_pos
df <- tibble(y, seronegative = (1 - pi_t) * f_neg,
             seropositive = pi_t * f_pos,
             mixture = mix) |>
  pivot_longer(-y)
ggplot(df, aes(y, value, colour = name, linetype = name)) +
  geom_line(linewidth = 1.4) +
  geom_vline(xintercept = log(5), linetype = "22", colour = "grey40") +
  annotate("text", x = log(5) + 0.05, y = max(mix) * 0.95,
           label = "LoD = ln(5)", hjust = 0, size = 4) +
  scale_colour_manual(values = c(seronegative = "#1f77b4",
                                 seropositive = "#d62728",
                                 mixture = "black")) +
  scale_linetype_manual(values = c(seronegative = "solid",
                                   seropositive = "solid",
                                   mixture = "22")) +
  labs(x = "log antibody titer (IU/mL)", y = "density",
       colour = NULL, linetype = NULL,
       title = "Two-component mixture, left-censored at LoD") +
  theme(legend.position = "bottom")
```
:::
::::

## Vaccine effectiveness is a hazard ratio {.smaller}

:::: {.columns}
::: {.column width="55%"}
$$VE = 1 - RR = 1 - \frac{\Pr(\text{infection} \mid V)}{\Pr(\text{infection} \mid \bar{V})}$$

But which $RR$? Cumulative, rate, or **hazard ratio**?

In survival language:
$$\lambda_V(t) = e^{-\beta(t)}\, \lambda_{\bar{V}}(t),\qquad VE(t) = 1 - e^{-\beta(t)}$$

- **Waning immunity:** $\beta(t)$ time-varying ⇒ a flexible Cox / Aalen model
- **Depletion of high-risk first** (frailty!) makes VE look smaller than it is
- **Time-varying covariate:** vaccination is a transition between states — multistate again

::: {.fragment}
*Case-study task 4 / 5:* breakthrough infections under vaccination + waning are a time-to-event problem with a time-dependent exposure.
:::
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
t <- seq(0, 200, 1)
# True VE wanes from 0.9 to 0.3 over 6 months
true_VE <- 0.3 + 0.6 * exp(-t / 90)
# Naive VE biased downward by frailty depletion (high-risk infect first)
naive_VE <- true_VE - 0.1 * (1 - exp(-t / 60))
df <- tibble(t, true = true_VE, `naive (frailty-biased)` = naive_VE) |>
  pivot_longer(-t)
ggplot(df, aes(t, value, colour = name)) +
  geom_line(linewidth = 1.5) +
  scale_colour_manual(values = c(true = "#1f77b4",
                                 `naive (frailty-biased)` = "#d62728")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(x = "days since vaccination", y = "VE(t)", colour = NULL,
       title = "Apparent VE drops faster than true VE",
       subtitle = "Selection of susceptibles + waning, combined") +
  theme(legend.position = "bottom")
```
:::
::::

## Excess mortality as competing risks {.smaller}

:::: {.columns}
::: {.column width="55%"}
The mortality dataset records *time of death* with **no cause attribution**.

The standard fix:

$$
\text{excess}(t) = \text{observed}(t) - E[\,\text{baseline mortality}\,](t)
$$

In survival language, this is just a **cause-specific** decomposition where the cause is unobserved:

- baseline hazard $\mu(a)$ estimated from pre-pandemic data
- excess hazard attributed to the epidemic
- competing-risks structure: $S \to D_{\text{epi}}$ vs $\cdot \to D_{\text{other}}$

::: {.fragment}
*Case-study task 5:* relate excess mortality to infection and vaccination histories — a multistate model with time-dependent covariates.
:::
:::
::: {.column width="45%"}
```{r}
#| fig-height: 4.7
set.seed(8)
weeks <- 1:60
baseline <- 1000 + 20 * sin(2 * pi * weeks / 52)
epi_signal <- 700 * dnorm(weeks, 35, 5) / max(dnorm(weeks, 35, 5))
observed <- baseline + epi_signal + rnorm(length(weeks), 0, 25)
df <- tibble(week = weeks, baseline, observed)
ggplot(df, aes(week)) +
  geom_ribbon(aes(ymin = baseline, ymax = observed),
              fill = "#fff2cc", alpha = 0.9) +
  geom_line(aes(y = observed), colour = "#d62728", linewidth = 1.4) +
  geom_line(aes(y = baseline), colour = "#1f77b4", linewidth = 1.4,
            linetype = "22") +
  annotate("text", x = 35, y = 1500, label = "excess",
           colour = "#5a4500", size = 5, fontface = "bold") +
  labs(x = "week", y = "weekly deaths",
       title = "Excess mortality = area between curves",
       subtitle = "Blue dashed: expected baseline. Red: observed.")
```
:::
::::

# Synthesis {background-color="#444444" style="color:white;"}

## The full dictionary {.smaller}

| Survival analysis | Infectious disease modelling |
|---|---|
| Hazard $h(t)$ | Force of infection $\lambda(t)$ |
| Cumulative hazard $H(t)$ | Cumulative incidence per susceptible |
| Survival $S(t)$ | Susceptible fraction |
| Frailty $Z$ | Heterogeneous susceptibility / activity |
| Cox PH | Proportional FoI / risk-stratified models |
| Multistate model | Compartmental model (SIR, SEIR, …) |
| Sojourn time | Latent / infectious period |
| Competing risks | Recovery vs. death; multiple pathogens |
| Right-truncation | Reporting delay / nowcasting |
| Doubly interval-censoring | Incubation / generation interval estimation |
| Restricted mean survival time | $R_0$ as $\int \beta(\tau)S_I(\tau) d\tau$ |
| Recurrent events | Reinfection / waning immunity |
| $1 - S(t)$ | Time-dependent seroprevalence $\pi(t)$ |
| Misclassification correction | Rogan–Gladen estimator |
| Time-dependent hazard ratio | Vaccine effectiveness $VE(t)$ |
| Excess hazard | Excess mortality |

## What infectious disease modelling adds {.smaller}

:::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- **Endogenous hazard.** $\lambda(t)$ depends on the population state.
- **Network structure.** Risk sets are not exchangeable — contacts matter.
- **Reinfection.** Recurrent-event survival, waning immunity as a hazard model.
- **Causal interventions at scale.** Vaccination, NPIs — population counterfactuals.
:::
:::
::: {.column width="50%"}
::: {.incremental}
- **Phylogenetics.** Trees of pathogens give independent windows on the same hazard process.
- **Spatial dependence.** Latent point processes; classical survival rarely needs them.
- **Real-time decisions.** Right-truncation matters every day, not only at study end.
:::
:::
::::

## Software bridges

| Need | Survival side | Epi side |
|---|---|---|
| Hazards, KM, Cox | `survival`, `survminer` | shared |
| Multistate / competing risks | `mstate`, `flexsurv` | `EpiModel`, `SimInf` |
| ODE / compartmental | — | `deSolve`, `pomp` |
| Rₜ estimation | — | `EpiEstim`, `EpiNow2`, `epinowcast` |
| Doubly interval-censored | `icenReg`, `Icens` | `coarseDataTools`, `epitrix` |
| Nowcasting | — | `nobBs`, `epinowcast` |

::: {.fragment}
Almost every "epi" package above is a thin layer over machinery your community wrote first.
:::

## Take-aways {.smaller}

::: {.incremental}
1. **Hazard = force of infection.** Same equation, two communities.
2. **SIR/SEIR are multistate models** with endogenous transition intensities.
3. **R₀ and Rₜ** are integrals and ratios of objects you already estimate.
4. **Censoring and truncation** in outbreak data are the same problems your field has solved for decades.
5. The remaining hard problems — networks, reinfection, real-time inference — are *exactly* where survival analysis can push the field forward.
:::

## Further reading {.smaller}

- Kermack & McKendrick (1927). *A contribution to the mathematical theory of epidemics.* **Proc. R. Soc. A.**
- Bernoulli (1760). *Essai d'une nouvelle analyse de la mortalité causée par la petite vérole…*
- Anderson & May (1991). *Infectious Diseases of Humans.* OUP.
- Andersson & Britton (2000). *Stochastic Epidemic Models and Their Statistical Analysis.* Springer.
- Held, Hens, O'Neill & Wallinga, eds. (2019). *Handbook of Infectious Disease Data Analysis.* CRC.
- Cori, Ferguson, Fraser & Cauchemez (2013). A new framework and software to estimate $R_t$. **AJE.**
- Brookmeyer & Gail (1988). A method for obtaining short-term projections of the AIDS epidemic. **JASA.**
- Keeling & Rohani (2008). *Modeling Infectious Diseases in Humans and Animals.* Princeton.
- Wallinga & Lipsitch (2007). How generation intervals shape the relationship between growth rates and reproductive numbers. **Proc. R. Soc. B.**
- Rogan & Gladen (1978). Estimating prevalence from the results of a screening test. **AJE.**
- Hens et al. (2012). *Modeling Infectious Disease Parameters Based on Serological and Social Contact Data.* Springer.
- Halloran et al. (2010). *Design and Analysis of Vaccine Studies.* Springer.

## Thank you {.center}

:::: {.columns}
::: {.column width="50%"}
**Discussion**

- Where does *your* method already apply?
- Where are the open methodological gaps?
- What would unlock collaboration?
:::
::: {.column width="50%"}
```{mermaid}
%%| fig-width: 6
flowchart LR
  SA(("survival<br/>analysis")):::sa <-->|"shared math"| ID(("infectious<br/>disease<br/>modelling")):::id
  classDef sa fill:#dfeefc,stroke:#1f77b4,color:#0b3d66
  classDef id fill:#fde2e2,stroke:#d62728,color:#5a0c0c
```
:::
::::
