---
title: "Infectious Disease Modelling"
subtitle: "An introduction for survival analysts"
author: "Kylie Ainslie"
date: today
format:
  revealjs:
    theme: simple
    slide-number: c/t
    incremental: false
    chalkboard: true
    transition: slide
    fig-align: center
    fig-width: 7
    fig-height: 3.4
    code-fold: false
    code-line-numbers: false
    smaller: true
    width: 1280
    height: 720
    margin: 0.04
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false
    dev: "svglite"
    fig.align: "center"
    fig.height: 3.4
    fig.width: 7
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(deSolve)
})
theme_set(theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold")))
pal <- c(S = "#1f77b4", E = "#9467bd", I = "#d62728", R = "#2ca02c", D = "#7f7f7f")
sir <- function(t, y, p) {
  with(as.list(c(y, p)), {
    dS <- -beta * S * I / N
    dI <-  beta * S * I / N - gamma * I
    dR <-  gamma * I
    list(c(dS, dI, dR))
  })
}
```

## The situation

:::: {.columns}
::: {.column width="60%"}
It's Monday morning. A new respiratory illness has been reported:

- **112 cases** last week
- **847 cases** this week

Hospitals want to know how many beds to prepare.
The government wants to know whether to close schools.
The press wants to know if this will get worse.

**You have one week of data. What do you do?**
:::
::: {.column width="40%"}
```{r}
df <- tibble(week = c("last week", "this week"),
             cases = c(112, 847),
             week = factor(week, levels = c("last week", "this week")))
ggplot(df, aes(week, cases, fill = week)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(cases)), vjust = -0.4,
            fontface = "bold", size = 5) +
  scale_fill_manual(values = c("#1f77b4", "#d62728")) +
  scale_y_continuous(limits = c(0, 1000)) +
  labs(x = NULL, y = "reported cases") +
  theme(axis.text.x = element_text(size = 13, face = "bold"))
```
:::
::::

## Three questions every epidemiologist faces

::: {.incremental}
1. **Describe** — What is the epidemic doing right now?
   *How many people are infected? Is it growing or shrinking?*

2. **Predict** — What will it do next?
   *How many hospitalisations next month? When will it peak?*

3. **Decide** — What should we do about it?
   *Do school closures help? How many people need to be vaccinated?*
:::

::: {.fragment}
> All three require a **model of how infection spreads through a population**.
:::

## Let's build one from scratch

We want to describe what happens to people during an epidemic.

::: {.fragment}
**Step 1: What states can a person be in?**
:::

::: {.fragment}
- **S**usceptible — not yet infected, could become infected
- **I**nfectious — currently infected and able to transmit
- **R**ecovered — immune, no longer at risk
:::

::: {.fragment}
That's it. Three states. Now: how do people move between them?
:::

## The SIR model

```{mermaid}
%%| fig-width: 9
flowchart LR
  S(["<b>S</b>usceptible\n<i>at risk</i>"]):::s -->|"becomes infected"| I(["<b>I</b>nfectious\n<i>can transmit</i>"]):::i
  I -->|"recovers"| R(["<b>R</b>ecovered\n<i>immune</i>"]):::r
  classDef s fill:#cfe2f3,stroke:#1f77b4,stroke-width:2px,color:#0b3d66
  classDef i fill:#fde2e2,stroke:#d62728,stroke-width:2px,color:#5a0c0c
  classDef r fill:#d9f1d9,stroke:#2ca02c,stroke-width:2px,color:#1b4a1b
```

::: {.fragment}
Two transitions. Two rates. Each rate is a **hazard**.
:::

::: {.fragment}
This diagram should look familiar.
:::

## You have seen this before

:::: {.columns}
::: {.column width="50%"}
**Survival analysis**

```{mermaid}
%%| fig-width: 4.5
flowchart LR
  H[Healthy]:::s -->|"h(t)"| D[Diseased]:::i
  D -->|"μ(t)"| E[Event / Death]:::r
  classDef s fill:#cfe2f3,stroke:#1f77b4,stroke-width:2px
  classDef i fill:#fde2e2,stroke:#d62728,stroke-width:2px
  classDef r fill:#d9f1d9,stroke:#2ca02c,stroke-width:2px
```
:::
::: {.column width="50%"}
**Infectious disease modelling**

```{mermaid}
%%| fig-width: 4.5
flowchart LR
  S[Susceptible]:::s -->|"λ(t)"| I[Infectious]:::i
  I -->|"γ"| R[Recovered]:::r
  classDef s fill:#cfe2f3,stroke:#1f77b4,stroke-width:2px
  classDef i fill:#fde2e2,stroke:#d62728,stroke-width:2px
  classDef r fill:#d9f1d9,stroke:#2ca02c,stroke-width:2px
```
:::
::::

::: {.fragment}
The SIR model **is** a multistate model. Same structure. Same mathematics. Different vocabulary.
:::

## But there is one critical difference

:::: {.columns}
::: {.column width="50%"}
**In a survival study**

The hazard $h(t)$ is determined by:

- the patient's age, treatment, covariates
- external factors

It does **not** depend on what other patients are doing.
:::
::: {.column width="50%"}
::: {.fragment}
**In an epidemic**

The hazard — the **force of infection** $\lambda(t)$ — is:

$$\lambda(t) = \beta \cdot \frac{I(t)}{N}$$

It depends on **how many people are infectious right now**.

Your risk today is created by the epidemic itself.
:::
:::
::::

::: {.fragment}
> The hazard is **endogenous**. It rises and falls with $I(t)$.
:::

## What does β · I(t)/N mean?

$$\lambda(t) = \underbrace{\beta}_{\text{transmission rate}} \cdot \underbrace{\frac{I(t)}{N}}_{\text{probability your contact is infectious}}$$

::: {.incremental}
- $\beta$ = (contacts per day) × (probability of transmission per contact)
- $I(t)/N$ = fraction of the population that is currently infectious
- $\lambda(t)$ = your personal hazard of infection on day $t$
:::

::: {.fragment}
As $I(t)$ grows, your hazard grows. As $I(t)$ falls, your hazard falls.
*This feedback loop is what makes epidemic dynamics different.*
:::

## The SIR equations

Putting the transition rates together:

$$\frac{dS}{dt} = -\underbrace{\beta \frac{I}{N}}_{\lambda(t)} \cdot S \qquad \frac{dI}{dt} = \beta \frac{SI}{N} - \gamma I \qquad \frac{dR}{dt} = \gamma I$$

::: {.incremental}
- $S$ is depleted at rate $\lambda(t) \cdot S$ — susceptibles getting infected
- $I$ grows from new infections, shrinks as people recover at rate $\gamma$
- The sojourn time in $I$ is $\text{Exp}(\gamma)$, with mean $1/\gamma$ days
:::

::: {.fragment}
These are just differential equations describing the multistate diagram from two slides ago.
:::

## What an epidemic looks like

```{r}
out <- as.data.frame(ode(c(S = 0.999, I = 0.001, R = 0),
                         seq(0, 120, 0.5), sir,
                         parms = c(beta = 0.4, gamma = 0.15, N = 1)))
out |>
  pivot_longer(-time, names_to = "state") |>
  mutate(state = factor(state, levels = c("S", "I", "R"))) |>
  ggplot(aes(time, value, colour = state)) +
  geom_line(linewidth = 1.6) +
  scale_colour_manual(values = pal,
                      labels = c(S = "Susceptible", I = "Infectious", R = "Recovered")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "day", y = "fraction of population", colour = NULL,
       title = "A single SIR epidemic (β = 0.4, γ = 0.15)") +
  theme(legend.position = "bottom")
```

::: {.fragment}
The epidemic rises, peaks, and falls — all driven by the depletion of susceptibles.
:::

# How fast does it spread? {background-color="#1f77b4" style="color:white;"}

## One number summarises it

**How many people does one infectious person infect, on average, before they recover?**

::: {.fragment}
That number is $R_0$ — the **basic reproduction number**.
:::

:::: {.columns}
::: {.column width="50%"}
::: {.fragment}
```{mermaid}
%%| fig-width: 5
flowchart TB
  A(("1")):::seed --> B1(("2")):::gen1
  A --> B2(("2")):::gen1
  A --> B3(("2")):::gen1
  B1 --> C1(("3")):::gen2
  B1 --> C2(("3")):::gen2
  B2 --> C3(("3")):::gen2
  B2 --> C4(("3")):::gen2
  B3 --> C5(("3")):::gen2
  B3 --> C6(("3")):::gen2
  classDef seed fill:#d62728,stroke:#5a0c0c,color:white
  classDef gen1 fill:#f7b7b7,stroke:#d62728
  classDef gen2 fill:#fde2e2,stroke:#d62728
```
:::
:::
::: {.column width="50%"}
::: {.incremental}
- If $R_0 > 1$ → each case generates more than one new case → epidemic grows
- If $R_0 < 1$ → epidemic fades out
- If $R_0 = 1$ → endemic equilibrium
:::
:::
::::

## R₀ in the SIR model

::: {.incremental}
- One infectious person is in state $I$ for an average of $1/\gamma$ days
- During that time they make contact at rate $\beta$
- So on average they infect $\beta \times 1/\gamma$ others in a fully susceptible population
:::

::: {.fragment}
$$R_0 = \frac{\beta}{\gamma}$$

A ratio of **two survival quantities**: the transmission rate and the recovery rate.
:::

::: {.fragment}
For the epidemic above: $R_0 = 0.4 / 0.15 \approx 2.7$. Each case generates ~2–3 secondary cases.
:::

## As the epidemic unfolds: Rₜ

$R_0$ assumes a fully susceptible population. As susceptibles are depleted, the effective reproduction number $R_t$ declines:

$$R_t = R_0 \cdot \frac{S(t)}{N}$$

```{r}
out <- as.data.frame(ode(c(S = 0.999, I = 0.001, R = 0),
                         seq(0, 120, 0.5), sir,
                         parms = c(beta = 0.4, gamma = 0.15, N = 1)))
out$Rt <- (0.4 / 0.15) * out$S
ggplot(out, aes(time, Rt)) +
  geom_hline(yintercept = 1, linetype = "22", colour = "grey40") +
  geom_line(linewidth = 1.6, colour = "#1f77b4") +
  annotate("text", x = 85, y = 1.12, label = "Rt = 1  (epidemic turning point)",
           colour = "grey30", size = 4, hjust = 0) +
  labs(x = "day", y = expression(R[t]),
       title = "Rt declines as susceptibles are depleted")
```

The epidemic turns when $R_t$ crosses 1 — the point at which $S(t)/N = 1/R_0$.

# The connection to survival analysis {background-color="#2ca02c" style="color:white;"}

## The hazard function, renamed

:::: {.columns}
::: {.column width="50%"}
**Survival analysis**

$$h(t) = \lim_{\Delta t \to 0} \frac{\Pr(t \le T < t+\Delta t \mid T \ge t)}{\Delta t}$$

$$S(t) = \exp\!\left\{-\int_0^t h(u)\, du\right\}$$

$S(t)$ = probability of not yet having the event
:::
::: {.column width="50%"}
::: {.fragment}
**Infectious disease modelling**

$$\lambda(t) = \beta \cdot \frac{I(t)}{N}$$

$$S(t) = \exp\!\left\{-\int_0^t \lambda(u)\, du\right\}$$

$S(t)$ = fraction of the population still susceptible
:::
:::
::::

::: {.fragment}
Same equation. $S(t)$ is simultaneously a survival curve and the susceptible fraction.
:::

## Adding a latent period: SEIR

Many infections have an incubation period — you are infected but not yet infectious.

```{mermaid}
%%| fig-width: 10
flowchart LR
  S(["S\n<i>susceptible</i>"]):::s -->|"λ(t)"| E(["E\n<i>exposed / latent</i>"]):::e
  E -->|"σ"| I(["I\n<i>infectious</i>"]):::i
  I -->|"γ"| R(["R\n<i>recovered</i>"]):::r
  classDef s fill:#cfe2f3,stroke:#1f77b4
  classDef e fill:#e8d8f7,stroke:#9467bd
  classDef i fill:#fde2e2,stroke:#d62728
  classDef r fill:#d9f1d9,stroke:#2ca02c
```

::: {.incremental}
- The sojourn in **E** is the incubation period — a waiting-time distribution
- Mean sojourn = $1/\sigma$ days; Exp($\sigma$) by default
- In the workshop case study, the underlying model is SIRVD — same idea, extra states
:::

## Competing risks: when infection can kill

```{mermaid}
%%| fig-width: 9
flowchart LR
  S(["S"]):::s -->|"λ(t)"| I(["I"]):::i
  I -->|"γ · (1 − IFR)"| R(["R  recovered"]):::r
  I -->|"γ · IFR"| D(["D  dead"]):::d
  classDef s fill:#cfe2f3,stroke:#1f77b4
  classDef i fill:#fde2e2,stroke:#d62728
  classDef r fill:#d9f1d9,stroke:#2ca02c
  classDef d fill:#eeeeee,stroke:#555
```

::: {.incremental}
- Two competing exits from $I$: recovery or death
- The **infection fatality ratio (IFR)** is the cause-specific cumulative incidence as $t \to \infty$
- The standard competing-risks toolkit — cause-specific hazards, Fine-Gray — applies directly
:::

## The vocabulary dictionary

| Survival analysis | Infectious disease modelling |
|:---|:---|
| Hazard $h(t)$ | Force of infection $\lambda(t)$ |
| Survival $S(t)$ | Susceptible fraction |
| Multistate model | Compartmental model (SIR, SEIR, …) |
| Sojourn time | Latent / infectious period |
| Competing risks | Recovery vs death |
| Frailty | Heterogeneous susceptibility |
| Right-censoring | Unconfirmed infection |
| Right-truncation | Reporting delay |
| Doubly interval-censored | Incubation period estimation |
| Time-varying hazard ratio | Vaccine effectiveness $VE(t)$ |
| Excess hazard | Excess mortality |

# The data challenges {background-color="#d62728" style="color:white;"}

## What data do we actually have?

```{mermaid}
%%| fig-width: 10
flowchart LR
  Inf["infection\n(unobserved)"]:::u --> Sym["symptom\nonset"]:::o
  Sym --> Rep["confirmed /\nreported"]:::o
  Rep --> Out["outcome\n(R or D)"]:::o
  classDef u fill:#eeeeee,stroke:#888,stroke-dasharray:4 4,color:#555
  classDef o fill:#d9f1d9,stroke:#2ca02c
```

::: {.incremental}
- Infection times are **almost never** observed
- We see: symptom onset, confirmation date, outcome
- Everything we care about involves at least one **censored or truncated** endpoint
- The problems are familiar — only the names are different
:::

## Reporting delays = right-truncation

Cases reported with a delay. If we condition on what has been reported **by today**, we systematically miss recent cases.

```{r}
set.seed(3)
days <- 0:80
mu <- 160 * dnorm(days, 50, 9) / max(dnorm(days, 50, 9))
true_inc <- rpois(length(days), pmax(mu, 0.4))
delay_pmf <- dgamma(0:30, shape = 2, rate = 0.4)
delay_pmf <- delay_pmf / sum(delay_pmf)
today <- 70
reported <- rep(0, length(days))
for (d in seq_along(days)) {
  for (k in 0:30) {
    if (days[d] + k <= today)
      reported[d] <- reported[d] + true_inc[d] * delay_pmf[k+1]
  }
}
df <- tibble(day = days, `true incidence` = true_inc, `reported by today` = reported) |>
  pivot_longer(-day)
ggplot(df, aes(day, value, fill = name)) +
  geom_col(data = filter(df, name == "true incidence"),
           fill = NA, colour = "#1f77b4", linewidth = 0.4) +
  geom_col(data = filter(df, name == "reported by today"),
           fill = "#d62728", alpha = 0.8) +
  geom_vline(xintercept = today, linetype = "22") +
  annotate("text", x = today + 1, y = max(true_inc) * 0.9,
           label = "today →", hjust = 0, size = 3.5) +
  labs(x = "day of symptom onset", y = "cases",
       title = "Right-truncated incidence",
       subtitle = "Outline: true incidence. Red: what we see today. Recent cases are undercounted.")
```

This is the same problem as **right-truncation** in follow-up studies. The fix — nowcasting — is a deconvolution of the delay distribution.

## Incubation period = doubly interval-censored

We know an exposure window $[E_L, E_R]$ and a symptom-onset window $[O_L, O_R]$.
The incubation period $T = O - E$ is **doubly interval-censored** — a standard survival problem.

```{r}
set.seed(11)
n <- 12
df <- tibble(
  case = factor(paste("case", seq_len(n))),
  EL = runif(n, -3, -1), ER = EL + runif(n, 0.5, 2),
  OL = ER + runif(n, 1.5, 6), OR = OL + runif(n, 0.3, 1.5)
)
ggplot(df) +
  geom_segment(aes(x = EL, xend = ER, y = case, yend = case),
               colour = "#1f77b4", linewidth = 3.5, lineend = "round") +
  geom_segment(aes(x = OL, xend = OR, y = case, yend = case),
               colour = "#d62728", linewidth = 3.5, lineend = "round") +
  labs(x = "days", y = NULL,
       title = "Exposure windows (blue) and onset windows (red)",
       subtitle = "Both endpoints censored. EM, Bayesian augmentation, or multiple imputation all apply.")
```

## Vaccine effectiveness is a hazard ratio

$$VE(t) = 1 - \frac{\lambda_{\text{vaccinated}}(t)}{\lambda_{\text{unvaccinated}}(t)} = 1 - HR(t)$$

```{r}
t <- seq(0, 180, 1)
true_VE  <- 0.3 + 0.6 * exp(-t / 90)
naive_VE <- true_VE - 0.12 * (1 - exp(-t / 50))
tibble(t, `true VE` = true_VE, `naive estimate (biased)` = naive_VE) |>
  pivot_longer(-t) |>
  ggplot(aes(t, value, colour = name)) +
  geom_line(linewidth = 1.5) +
  scale_colour_manual(values = c(`true VE` = "#1f77b4",
                                 `naive estimate (biased)` = "#d62728")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(x = "days since vaccination", y = "VE(t)", colour = NULL,
       title = "Apparent VE drops faster than true VE",
       subtitle = "Depletion of high-risk susceptibles creates the illusion of waning") +
  theme(legend.position = "bottom")
```

::: {.fragment}
The downward bias comes from **frailty**: high-risk unvaccinated individuals are infected first, leaving a progressively lower-risk comparator group — the same artefact as frailty-induced declining hazards in survival data.
:::

# The workshop {background-color="#444444" style="color:white;"}

## Your case study

A simulated population-wide outbreak, generated by an individual-based model (**HERMES**).

:::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- N = 1,000,000 individuals in households, schools, workplaces
- SIRVD structure with heterogeneity in contacts and susceptibility
- Lockdown at day 20 → exit at day 35 → vaccination from day 45
- You do **not** see the full dataset until Day 5
:::
:::
::: {.column width="45%"}
**Available datasets**

| | |
|--|--|
| `final_demographic_data` | age, sex, household |
| `final_incidence_data` | reported cases + hospitalisations |
| `final_sero_data` | IgG titres at days 10, 20, 30, 40 |
| `final_contact_tracing_data` | infector–infectee pairs |
| `final_mortality_data` | times of death |
| `final_vaccination_data` | vaccination times |
:::
::::

## Five questions — one per group

| Group | Question |
|:---:|:---|
| **A** | Key epidemiological parameters: R₀, Rₜ, incubation period, reporting delay |
| **B** | Serial interval distribution from contact tracing data |
| **C** | Reporting delays and underreporting, using serological data |
| **D** | Vaccine effectiveness and the role of susceptible depletion |
| **E** | Excess mortality and its relation to infection and vaccination |

::: {.fragment}
Every one of these is a survival analysis problem in an epidemic setting.
:::

## Your toolkit applies

| Workshop question | SA methods you'll reach for |
|:---|:---|
| **A** Epi parameters | Parametric survival, renewal equation |
| **B** Serial interval | Censored time-to-event, interval censoring |
| **C** Underreporting | Mixture models, truncation correction |
| **D** Vaccine effectiveness | Time-varying covariates, Cox / Aalen |
| **E** Excess mortality | Competing risks, relative survival |

::: {.fragment}
> The goal this week is not to learn new methods. It is to see where your methods already live inside IDM — and where the gaps are.
:::

## Thank you {.center}

:::: {.columns}
::: {.column width="55%"}
**Questions for discussion**

- Where does your own work already connect to IDM?
- Where do the methods diverge?
- What would help the most this week?
:::
::: {.column width="45%"}
```{mermaid}
%%| fig-width: 5
flowchart LR
  SA(("survival\nanalysis")):::sa <-->|"shared\nmath"| ID(("infectious\ndisease\nmodelling")):::id
  classDef sa fill:#dfeefc,stroke:#1f77b4,color:#0b3d66,stroke-width:2px
  classDef id fill:#fde2e2,stroke:#d62728,color:#5a0c0c,stroke-width:2px
```
:::
::::

---

# Appendix {background-color="#888888" style="color:white;"}

## The serial interval vs generation interval {.smaller}

:::: {.columns}
::: {.column width="55%"}
- **Generation interval:** time between *infection* of infector and *infection* of infectee — governs the renewal equation
- **Serial interval:** time between *symptom onset* of infector and *symptom onset* of infectee — what we observe in line lists

$$\text{SI} = \text{GI} + (\text{incubation}_{\text{infectee}} - \text{incubation}_{\text{infector}})$$

Same mean; SI has larger variance and can be **negative**.

In the contact tracing data (Group B): onset times always observed, infection times sometimes interval-censored.
:::
::: {.column width="45%"}
```{r}
set.seed(9)
n <- 5000
GI <- rgamma(n, shape = 4, rate = 4/5.2)
inc_or <- rgamma(n, shape = 4, rate = 4/4)
inc_ee <- rgamma(n, shape = 4, rate = 4/4)
SI <- GI + (inc_ee - inc_or)
tibble(value = c(GI, SI),
       kind  = rep(c("Generation interval", "Serial interval"), each = n)) |>
  ggplot(aes(value, fill = kind, colour = kind)) +
  geom_density(alpha = 0.35, linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "22") +
  scale_fill_manual(values  = c("Generation interval" = "#d62728",
                                "Serial interval"    = "#1f77b4")) +
  scale_colour_manual(values = c("Generation interval" = "#d62728",
                                 "Serial interval"    = "#1f77b4")) +
  labs(x = "days", y = "density", fill = NULL, colour = NULL,
       title = "Same mean — but SI can be negative") +
  theme(legend.position = "bottom")
```
:::
::::

## Heterogeneity and susceptible depletion {.smaller}

:::: {.columns}
::: {.column width="50%"}
Not everyone is equally susceptible. High-contact or high-risk individuals are infected first — depleting from the susceptible pool.

$$\lambda_i(t) = c_i \cdot \beta \cdot \frac{I(t)}{N}$$

The $c_i$ are frailties. The population-average FoI **declines over time** even without any intervention — because the remaining susceptibles are lower-risk.

This creates an apparent decline in vaccine effectiveness even when true protection hasn't waned (Group D).
:::
::: {.column width="50%"}
```{r}
t <- seq(0, 12, 0.05)
groups <- tibble(g = c("low risk (c=0.4)", "average (c=1)", "high risk (c=2.5)"), c = c(0.4, 1, 2.5))
df <- expand.grid(t = t, g = groups$g) |> left_join(groups, by = "g") |>
  mutate(S = exp(-0.25 * c * t))
pop <- df |> group_by(t) |> summarise(S = mean(S), .groups = "drop") |> mutate(g = "population mean")
ggplot() +
  geom_line(data = df, aes(t, S, colour = g), linewidth = 1.2) +
  geom_line(data = pop, aes(t, S), colour = "black", linewidth = 1.6, linetype = "22") +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "time", y = "S(t)", colour = NULL,
       title = "High-risk individuals deplete first",
       subtitle = "Black: population mean. Apparent FoI declines without any intervention.") +
  theme(legend.position = "bottom")
```
:::
::::

## Excess mortality {.smaller}

:::: {.columns}
::: {.column width="55%"}
The mortality data records time of death — **no cause attribution**.

$$\text{excess}(t) = \text{observed}(t) - \mathbb{E}[\text{baseline mortality}](t)$$

In survival terms: decompose total hazard into epidemic and non-epidemic components. The epidemic component is the excess hazard.

- Baseline $\mu(a)$ estimated from pre-pandemic data
- Competing risks: $\to D_{\text{epidemic}}$ vs $\to D_{\text{other}}$
- Group E: relate excess mortality to infection and vaccination histories via multistate model with time-dependent covariates
:::
::: {.column width="45%"}
```{r}
set.seed(8)
weeks <- 1:60
baseline <- 1000 + 20 * sin(2 * pi * weeks / 52)
epi_signal <- 700 * dnorm(weeks, 35, 5) / max(dnorm(weeks, 35, 5))
observed <- baseline + epi_signal + rnorm(length(weeks), 0, 25)
tibble(week = weeks, baseline, observed) |>
  ggplot(aes(week)) +
  geom_ribbon(aes(ymin = baseline, ymax = observed), fill = "#fff2cc", alpha = 0.9) +
  geom_line(aes(y = observed), colour = "#d62728", linewidth = 1.4) +
  geom_line(aes(y = baseline), colour = "#1f77b4", linewidth = 1.4, linetype = "22") +
  annotate("text", x = 35, y = 1500, label = "excess", colour = "#5a4500",
           size = 4.5, fontface = "bold") +
  labs(x = "week", y = "weekly deaths",
       title = "Excess mortality = area between curves",
       subtitle = "Blue dashed: expected baseline. Red: observed.")
```
:::
::::

## Software {.smaller}

| Need | Survival | Epidemiology |
|:---|:---|:---|
| Hazards, KM, Cox | `survival`, `survminer` | (shared) |
| Multistate / competing risks | `mstate`, `flexsurv` | `EpiModel` |
| ODE / compartmental models | — | `deSolve`, `odin` |
| Rₜ estimation | — | `EpiEstim`, `EpiNow2` |
| Doubly interval-censored | `icenReg` | `coarseDataTools` |
| Nowcasting | — | `epinowcast` |
| Mixture models | `flexmix` | (shared) |

## Further reading

- Kermack & McKendrick (1927). A contribution to the mathematical theory of epidemics. *Proc. R. Soc. A.*
- Keeling & Rohani (2008). *Modeling Infectious Diseases in Humans and Animals.* Princeton.
- Grassly & Fraser (2008). Mathematical models of infectious disease transmission. *Nature Reviews Microbiology.*
- Cori et al. (2013). A new framework and software to estimate time-varying reproduction numbers. *American Journal of Epidemiology.*
- Lipsitch (2019). Challenges of vaccine effectiveness and waning studies. *Clinical Infectious Diseases.*
- Held, Hens, O'Neill & Wallinga, eds (2019). *Handbook of Infectious Disease Data Analysis.* CRC.
